Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-28 analysis/paper/paper.Rmd
KNIT BEFORE UPLOADING RPTS DRAFT

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June. Empty areas of the plot represent missing data that were not successfully transmitted from the tag"}
d <- dat.sf %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt))

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
_wind:temperature_ interaction [@udevitz2009cjz]), and day-of-year and time of day
interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
seal models included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of age-sex
classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial range and
we were interested in possible differences in the timing of haul-out behavior along a
latitudinal gradient. We omitted the latitudinal effect from ribbon and spotted seal
models since, during the spring, these species are most prevalent near the southern ice
edge in the Bering Sea [@conn2014mee].

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counterintuitive
when modeling the haul-out behavior of seal species with such a close association to sea
ice; seals haul out in the presence of sea ice and we could assess the local concentration
of sea ice during these events. However, seals also spend time in the water away from sea
ice. Thus, the challenge is how best to associate that seal's at-sea behavior with a sea
ice concentration. Would it be the concentration at the last known haul-out event? The
next known haul-out event? Or, the closest sea ice concentration available? Without a
meaningful means to associate a sea ice covariate with the seal throughout the duration of
the observation record, such parameters are not viable for inclusion in this model.

@@

For the second set of models examining inter-annual variation in sea ice use, we fitted
models with year by day-of-year interactions. However, in this case we only included
_year:day_ and _year:day2_, omitting the main effects of year as well as
_year:day3_ interactions because models with the latter effects were numerically
unstable. However, the modeled interactions were sufficient to allow shifts in haul-out
distribution, as one can show mathematically that a simple horizontal shift in timing of
haul-out distributions does not affect the main effects or cubic terms in a polynomial
regression model.

@@

spotted_type3 <- tibble(
  fit_spotted$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_type3 <- setNames(split(spotted_type3, 
                               seq(nrow(spotted_type3))), 
                         rownames(spotted_type3))
bearded_type3 <- tibble(
  fit_bearded$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
bearded_type3 <- setNames(split(bearded_type3, 
                               seq(nrow(bearded_type3))), 
                         rownames(bearded_type3))
```

@@

Spotted seal haul-out behavior appears most strongly influenced by barometric pressure
($F_{`r paste(spotted_type3$pressure$Num.df, spotted_type3$pressure$Den.df, sep=",")`}$ =
`r spotted_type3$pressure$F.value`; $p$ = `r spotted_type3$pressure$Prob.F`) and wind
($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$ =
`r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`) with temperature
having a moderate influence
($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$ =
`r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`). Spotted seals were
less likely to haul out at higher winds and more likely to be on the ice when temperatures
were relatively warm. Wind chill (_temperature:wind_) and precipitation were not as
influential. Differences in the magnitude of response between the age-sex classes is
present and consistent across each of the weather covariates.

@@

We also investigated the influence of environmental covariates on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind chill.
These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]) and a few
select studies of ice-associated seals [@hamilton2018, @perry2017]. Ribbon seals seemed to
be the most influenced by weather, with wind, temperature, barometric pressure, and
precipitation all being important components of the model. Spotted seals were most
affected by wind and barometric pressures. And, for bearded seals, the model indicated
wind and temperature had the greatest impact. In general, and as might be expected, seals
were more likely to haul-out when daily temperatures were warmer, winds speeds were less,
barometric pressure was high, and lower rates of precipitation. These weather conditions
are likely most beneficial for energetic considerations and may also provide the best
situation for predator detection. These results highlight the importance for haul-out
behavior models (and subsequent calculations of availability during surveys) to
incorporate weather covariates.

@@

We focused this paper on haul-out behavior of bearded, ribbon, and spotted seals. Ringed
seals are also present in the Bering and Chukchi Seas but exhibit qualitatively different
haul-out behavior. For instance, many ringed seals build subnivean layers in the snow on
top of the sea ice, which they frequently inhabit before snow melt causes their lairs to
collapse. Thus, the wet-dry sensor on a satellite tag could indicate that an animal is
hauled out, but it could still be unavailable to be detected because it is in a lair. We
hope to address availability of ringed seals using data from satellite tags, replicate
survey tracks, and auxiliary information about snow depth and timing of melt in a future
study.

>>>>

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, May, and June. Empty areas of the plot represent missing data that were not successfully transmitted from the tag."}
d <- dat.sf %>% 
  filter(speno == "HF2016_1002") %>% 
  dplyr::select(speno, haulout_dt, percent_dry) %>% 
  arrange(haulout_dt) %>% st_drop_geometry() %>% 
  mutate(year = lubridate::year(haulout_dt),
         month = lubridate::month(haulout_dt,label=TRUE),
         day = lubridate::day(haulout_dt),
         hour = lubridate::hour(haulout_dt))

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill (represented by a
*wind:temperature* interaction [@udevitz2009cjz]), and day-of-year and time of day
interactions to permit diurnal patterns to change throughout the year. Spotted and ribbon
seal models included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of age-sex
classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial range and
we were interested in possible differences in the timing of haul-out behavior along a
latitudinal gradient. We omitted the latitudinal effect from ribbon and spotted seal
models since, during the spring, these species are most prevalent near the southern ice
edge in the Bering Sea [@conn2014mee].

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem counterintuitive
when modeling the haul-out behavior of seal species with such a close association to sea
ice; seals haul out in the presence of sea ice, and we could assess the local
concentration of sea ice during these events (see [@olnes2020]). However, a major focus of
this study is to develop models applicable for aerial survey correction factors and using
sea ice as a covariate would almost certainly bias haul-out predictions towards those
seals that are on or near ice and therefore more likely to haul out.  Since aerial surveys
can only detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of the
bio-logger deployments used in this study were single attachments to the rear flipper of
the seal and do not provide at-sea locations, limiting our ability to fully evaluate
habitat preferences related to sea-ice. It should be noted that the aerial survey counts
are modeled with sea-ice and other habitat parameters. So, any final abundance estimates
that combine models of survey counts and our model for proportion of seals hauled out on
ice do account for the strong association between these species and sea-ice in the Bering
and Chukchi seas. Lastly, our study was limited to the spring season when seal haul-out
tendencies are strongly influenced by pupping, nursing, breeding, and molt and these
drivers are likely more influential than specific sea-ice concentration. Crawford et al
[-@crawford2019] compared haul-out probability models for bearded seals and found those
that only included season (and not sea-ice concentration) where the most parsimonious. 
For these reasons, we have elected not to use sea ice concentration as a predictor for
haul-out probability; studies to evaluate habitat selection in relation to sea-ice
parameters and other environmental covariates are in progress.

@@

For the second set of models examining inter-annual variation in sea ice use, we fitted
models with year by day-of-year interactions. However, in this case we only included
*year:day* and *year:day2*, omitting the main effects of year as well as *year:day3*
interactions because models with the latter effects were numerically unstable. However,
the modeled interactions were sufficient to allow shifts in haul-out distribution, as one
can show mathematically that a simple horizontal shift in timing of haul-out distributions
does not affect the main effects or cubic terms in a polynomial regression model.

@@

spotted_type3 <- tibble(
  fit_spotted$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_type3 <- setNames(split(spotted_type3, 
                               seq(nrow(spotted_type3))), 
                         rownames(spotted_type3))

bearded_type3 <- tibble(
  fit_bearded$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
bearded_type3 <- setNames(split(bearded_type3, 
                               seq(nrow(bearded_type3))), 
                         rownames(bearded_type3))
```

@@

Spotted seal haul-out behavior appears most strongly influenced by barometric pressure
($F_{`r paste(spotted_type3$pressure$Num.df, spotted_type3$pressure$Den.df, sep=",")`}$ =
`r spotted_type3$pressure$F.value`; $p$ = `r spotted_type3$pressure$Prob.F`) and wind
($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$ =
`r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`) with temperature
having a moderate influence
($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$ =
`r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`). Spotted seals were
less likely to haul out at higher winds and more likely to be on the ice when temperatures
were relatively warm. Wind chill (*temperature:wind*) and precipitation were not as
influential. Differences in the magnitude of response between the age-sex classes is
present and consistent across each of the weather covariates.

@@

We also investigated the influence of environmental covariates on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind chill.
These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]) and a few
select studies of ice-associated seals [@hamilton2018, @perry2017]. Ribbon seals seemed to
be the most influenced by weather, with wind, temperature, barometric pressure, and
precipitation all being important components of the model. Spotted seals were most
affected by wind and barometric pressures. And, for bearded seals, the model indicated
wind and temperature had the greatest impact. In general, and as might be expected, seals
were more likely to haul-out when daily temperatures were warmer, winds speeds were less,
barometric pressure was high, and rates of precipitation were. These weather conditions
are likely most beneficial for energetic considerations and may also provide the best
situation for predator detection. These results highlight the importance for haul-out
behavior models (and subsequent calculations of availability during surveys) to
incorporate weather covariates.

@@

We focused this paper on haul-out behavior of bearded, ribbon, and spotted seals. Ringed
seals are also present in the Bering and Chukchi Seas but exhibit qualitatively different
haul-out behavior. For instance, many ringed seals build subnivean layers in the snow on
top of the sea ice, which they frequently inhabit before snow melt causes their lairs to
collapse [@kelly1990]. Thus, the wet-dry sensor on a satellite tag could indicate that an
animal is hauled out, but it could still be unavailable to be detected because it is in a
lair. We hope to address availability of ringed seals using data from satellite tags,
replicate survey tracks, and auxiliary information about snow depth and timing of melt in
a future study.
